---
title: "Variant Annotation Using Bioconductor"
author: "Ana-Marija Krizanac"
date: "2026-03-10"
output: html_document
---

### Setup for Workshop

```{r}
# Install BiocManager (if not already installed)
install.packages("BiocManager")
```

```{r}
# Check Bioconductor version
BiocManager::version()
```

```{r}
# If needed, install a compatible version (should be 3.20–3.22)
BiocManager::install(version = "3.22")
```

```{r}
# Install required Bioconductor packages
BiocManager::install(c(
  "VariantAnnotation",
  "GenomicRanges",
  "rtracklayer",
  "biomaRt",
  "GenomeInfoDb"
))
```

```{r}
# Install required CRAN packages
install.packages(c(
  "dplyr",
  "tidyr",
  "ggplot2"
))
```

### Workshop

## Load required packages
```{r}
library(VariantAnnotation)
library(GenomicRanges)
library(rtracklayer)
library(biomaRt)
library(dplyr)
library(tidyr)
library(ggplot2)
library(GenomeInfoDb)
```

## Downloading and inspecting the data

# Load chr19:45,200,000–45,250,000 (Byrska-Bishop et al. Cell, 2022), http://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/working/20220422_3202_phased_SNV_INDEL_SV/
# VCF file is stored in a compressed manner (vcf.gz)
```{r}
vcf_file <- "https://raw.githubusercontent.com/ana123136/Variant-Annotation-Using-Bioconductor/main/1kGP_high_coverage_Illumina.chr19.filtered_50kb.vcf.gz"
index_file <- "https://raw.githubusercontent.com/ana123136/Variant-Annotation-Using-Bioconductor/main/1kGP_high_coverage_Illumina.chr19.filtered_50kb.vcf.gz.tbi"
```

```{r}
download.file(vcf_file, destfile = "1kGP_high_coverage_Illumina.chr19.filtered_50kb.vcf.gz", mode = "wb")
download.file(index_file, destfile = "1kGP_high_coverage_Illumina.chr19.filtered_50kb.vcf.gz.tbi", mode = "wb")
```

# Read the VCF
# library(VariantAnnotation) is needed for this
```{r}
vcf <- readVcf("1kGP_high_coverage_Illumina.chr19.filtered_50kb.vcf.gz", "hg38")
```

# Inspect file
```{r}
vcf
info(header(vcf))
header(vcf)
meta(header(vcf))
meta(header(vcf))$fileformat
meta(header(vcf))$contig
geno(vcf)$GT[1:2, ]
```

# Check file class 
```{r}
class(vcf)
```

# Extract genomic ranges (variant positions) from the VCF object
# library(GenomicRanges) is needed for this
```{r}
gr_variants <- rowRanges(vcf)
```

# Check file class
```{r}
class(gr_variants)
```

# Check all available conventions for chromosome names
# library(GenomeInfoDb) is needed for this
```{r}
styles <- genomeStyles("Homo_sapiens")
styles[ , (ncol(styles)-3):ncol(styles)]
```

# Check the naming convention for chromosome names in the file
```{r}
seqlevelsStyle(gr_variants)
```

# Check the genome annotation convention from the file
```{r}
genome(gr_variants)
```

# Setting genome annotation to correspond to Ensembl/NCBI-style name
```{r}
genome(gr_variants) <- "GRCh38"
```

# Setting chromosomal IDs to correspond to Ensembl/NCBI-style name ("chr19" --> "19")
```{r}
seqlevelsStyle(gr_variants) <- "NCBI"
```

# The entire file was read and not just a subset by genomic ranges, so we can ignore this column
```{r}
mcols(gr_variants)$paramRangeID <- NULL
```

# Inspect Granges VCF object
```{r}
colnames(mcols(gr_variants))
head(mcols(gr_variants))
table(seqnames(gr_variants))
range(gr_variants)
```

## Variant annotation – classification by size/type

# Create a list of variant types
# library(VariantAnnotation) is needed for this
```{r}
variant_types <- list(SNV = isSNV, Insertion = isInsertion, Deletion = isDeletion)
# isSNV: REF and ALT alleles are both a single nucleotide long
# isInsertion: REF allele is a single nucleotide, and the ALT allele is longer than a single nucleotide
# isDeletion: ALT allele is a single nucleotide, and the reference allele is longer than a single nucleotide
```

# Count SNVs, Insertions, Deletions
```{r}
counts <- sapply(names(variant_types), function(type) {
  cat("\n====================\nVariant type:", type, "\n")
  
  idx <- variant_types[[type]](vcf, singleAltOnly = TRUE) # TRUE --> only variants with a single alternate allele are evaluated
  print(table(idx))
  
  if (any(idx)) print(rowRanges(vcf[idx])[1:min(5, sum(idx))])
  
  sum(idx)
})
```

**Variant types plot**

# Plot variant types (classification of variants by their size)
```{r}
barplot(
  counts,
  col = c("darkblue", "deeppink", "gray"),
  main = "Variant Type Distribution",
  ylab = "Number of Variants",
  las = 2
)
```

# Check distribution of variant lengths
# library(GenomicRanges) is needed for this
```{r}
w <- width(gr_variants)
```

```{r}
table(w)
```

# Count the number of large variants
```{r}
large_variants <- gr_variants[w > 50]
length(large_variants)
```

# Check some of the large variants by inspecting metadata columns
```{r}
mcols(large_variants)
```

# ID field was stored as the row names of the GRanges object, not as a metadata column
```{r}
#mcols(gr_variants)$rsid <- rownames(mcols(gr_variants))
mcols(gr_variants)$rsid <- names(gr_variants)
```

# Count how many times each rsID appears in gr_variants
```{r}
rsid_counts <- table(mcols(gr_variants)$rsid)
```

# Identify multi-allelic rsIDs (rsID appears more than once)
```{r}
multi_rsid <- names(rsid_counts[rsid_counts > 1])
```

# Number of multi-allelic rsIDs
```{r}
length(multi_rsid)
```

# Check how many rows each multi-allelic rsID has (i.e., number of ALT alleles represented)
```{r}
multi_rsid_table <- data.frame(
  rsID = multi_rsid,
  n_rows = as.integer(rsid_counts[multi_rsid])
)
```

```{r}
multi_rsid_table # multi-allelic variants in this vcf file are presented as bi-allelic, i.e., ALT alleles are separated in different rows
```

# Check how one of the multi-allelic variants looks
```{r}
gr_variants[mcols(gr_variants)$rsid == "rs1320105362", c("REF", "ALT")]
# format of this variant when not presented as bi-allelic: "chr19  45239590  rs1320105362  G  A,T"
```

## Variant annotation – consequence type and associated phenotype

# Check available databases
# library(biomaRt) is needed for this
```{r}
listMarts(host = "https://www.ensembl.org")      
```

# Set up a connection to the Ensembl Variation database
```{r}
snp_mart <- useMart(
  biomart = "ENSEMBL_MART_SNP", # BioMart dedicated to SNPs
  dataset = "hsapiens_snp",     # SNPs from Homo sapiens dataset
  host    = "https://www.ensembl.org" # Ensembl server URL
)
```

# Extract unique non-missing rsIDs from the GRanges object
# These will be used to query Ensembl for annotations
```{r}
variant_list <- unique(na.omit(mcols(gr_variants)$rsid))
```

# Query Ensembl BioMart to retrieve functional consequences and associated phenotype descriptions for each rsID
```{r}
variant_annotations <- getBM(
  attributes = c(           # vector of attributes we want to retrieve
    "refsnp_id",            # rsID (dbSNP identifier)
    "chr_name",             # chromosome name
    "chrom_start",          # genomic start position
    "chrom_end",            # genomic end position
    "consequence_type_tv",  # variant consequence (transcript-based)
    "phenotype_description" # associated phenotype(s), if available
  ),
  filters = "snp_filter", # filter type: filter by SNP rsID
  values = variant_list,  # list of rsIDs to query
  mart = snp_mart         # BioMart connection object
)
```

```{r}
# Optional: reload previously saved annotations instead of querying BioMart
#annotat_connection <- url("https://raw.githubusercontent.com/ana123136/Variant-Annotation-Using-Bioconducto#r/main/variant_annotations_backup_consequences_and_phenotypes.rds", "rb")
#variant_annotations <- readRDS(annotat_connection)
#close(annotat_connection)
```

**Variant consequences plot**

```{r}
# Convert empty strings "" into NA
# library(tidyr) is needed for this
variant_annotations <- variant_annotations %>%
  mutate(
    consequence_type_tv = na_if(consequence_type_tv, ""),
    phenotype_description = na_if(phenotype_description, "")
  )
```

# Count consequence types
```{r}
consequence_counts <- variant_annotations %>%
  filter(!is.na(consequence_type_tv)) %>% # keep only rows where consequence_type_tv is not missing
  count(consequence_type_tv) %>% # counts the number of occurrences for each consequence type
  mutate(pct = n / sum(n) * 100) # calculates the percentage of each consequence type
```

# Create variants consequences plot
# library(ggplot2) is needed for this
```{r}
ggplot(consequence_counts, aes(x = "", y = pct, fill = consequence_type_tv)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Variant Consequences", fill = "Consequence Type") +
  theme_void() +
  theme(legend.position = "right")
```

**Top 20 phenotypes bar plot**

# Count phenotypes associated with variants
```{r}
phenotype_counts <- variant_annotations %>%
  filter(!is.na(phenotype_description)) %>% # removes rows where phenotype_description is missing
  count(phenotype_description) %>% # counts how many times each unique phenotype appears
  arrange(desc(n)) %>% # sorts the table in descending order
  slice_head(n = 20) # focus on the 20 most common phenotypes
```

# Create top 20 phenotypes bar plot
```{r}
ggplot(phenotype_counts, aes(x = reorder(phenotype_description, n), y = n, fill = phenotype_description)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 20 Phenotypes by SNP Count", x = "Phenotype", y = "Number of SNPs") +
  theme_minimal()
# Phenotype descriptions come from different databases, integrating different studies (GWAS Catalog, ClinVar)
# More than one phenotype possible for each variant
```

**Variant consequences & phenotypes heatmap**

# Create counts for consequence-phenotype combinations
```{r}
combo_counts <- variant_annotations %>%
  filter(!is.na(consequence_type_tv) & !is.na(phenotype_description)) %>% # filters out rows where either consequence_type_tv or phenotype_description is NA
  count(consequence_type_tv, phenotype_description)
```

# Keep only top 10 phenotypes
```{r}
top_phenos <- combo_counts %>%
  group_by(phenotype_description) %>%
  summarise(total = sum(n), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = 10) %>%
  pull(phenotype_description)
```

# Filter original combo_counts to only top phenotypes
```{r}
combo_counts_filtered <- combo_counts %>%
  filter(phenotype_description %in% top_phenos)
```

# Create heatmap of variant consequences & phenotypes
```{r}
ggplot(combo_counts_filtered, aes(x = consequence_type_tv, y = phenotype_description, fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightpink", high = "deeppink") +
  labs(
    title = "Consequences vs Phenotypes",
    x = "Variant Consequence",
    y = "Phenotype",
    fill = "SNP count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )
```

## Variant annotation – closest genes

# Map variants' positions to the closest genes

# Set up a connection to the Ensembl Genes database
```{r}
gene_mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  host = "https://www.ensembl.org"
)
```

# Query Ensembl BioMart to identify genes that overlap the genomic positions of the variants
```{r}
variant_annotations_genes <- getBM(
  attributes = c(
    "chromosome_name",
    "start_position",
    "end_position",
    "external_gene_name",
    "ensembl_gene_id",
    "description",
    "gene_biotype"
  ),
  filters = c("chromosome_name", "start", "end"),
  values = list(
    chromosome_name = variant_annotations$chr_name,
    start = variant_annotations$chrom_start,
    end   = variant_annotations$chrom_end
  ),
  mart = gene_mart
)
```

# Check identified genes
```{r}
head(variant_annotations_genes)
```

```{r}
# Optional: reload previously saved annotations instead of querying BioMart
#gene_annotat_connection <- url("https://raw.githubusercontent.com/ana123136/Variant-Annotation-Using-Bioconductor/main/variant_annotations_backup_genes.rds", "rb")
#variant_annotations_genes <- readRDS(gene_annotat_connection)
#close(gene_annotat_connection)
```

**Gene and variants locations plot**

# Make sure variants have a column to match genes
```{r}
variant_with_gene <- variant_annotations %>%
  left_join(
    variant_annotations_genes %>% 
      select(chromosome_name, start_position, end_position, external_gene_name),
    by = c("chr_name" = "chromosome_name"),
    relationship = "many-to-many"
  ) %>%
  filter(chrom_start >= start_position & chrom_start <= end_position)
```

# Visualize gene locations (blue) along the genome and variants located within those genes (pink)
```{r}
ggplot() +
  geom_segment(
    data = variant_annotations_genes,
    aes(
      x = start_position,
      xend = end_position,
      y = external_gene_name,
      yend = external_gene_name
    ),
    size = 4,
    color = "darkblue"
  ) +
  geom_point(
    data = variant_with_gene,
    aes(
      x = chrom_start,
      y = external_gene_name 
    ),
    color = "deeppink",
    size = 2
  ) +
  labs(
    x = "Genomic Position",
    y = "Gene",
    title = "Gene Locations with Variants"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))
```


```{r}
# Optional: reload previously saved annotations instead of querying BioMart
#variant_annotations_genes <- readRDS("variant_annotations_backup.rds")

## Optional: gene enrichment analysis
#BiocManager::install("clusterProfiler")
#library(clusterProfiler)
#BiocManager::install("org.Hs.eg.db")
#library(org.Hs.eg.db)
```


```{r}
# Convert Ensembl IDs to Entrez IDs
#entrez_ids <- bitr(variant_annotations_genes$ensembl_gene_id, fromType="ENSEMBL", toType="ENTREZID", OrgDb="org.Hs.eg.db")

# GO (Gene Ontology) enrichment analysis
#GO <- enrichGO(
#gene = entrez_ids$ENTREZID,
#OrgDb = org.Hs.eg.db,
#ont = "BP",  # Biological Process
#pAdjustMethod = "BH",
#qvalueCutoff = 0.05,
#readable = TRUE
#)

#dotplot(GO, showCategory = 10) + theme_minimal()
```

## Variant annotation – pathogenicity prediction

# Predicts effects of coding variants on protein function

# Keep only coding variants
# Filter coding variants using the correct column
```{r}
coding_variants <- variant_annotations %>%
  filter(grepl("missense|stop_gained|synonymous|frameshift|splice", 
               consequence_type_tv, ignore.case = TRUE))
```

# Check how many coding variants have rsIDs (refsnp_id)
```{r}
sum(!is.na(coding_variants$refsnp_id))
```

# Keep only variants with rsIDs for SIFT
```{r}
coding_rsids <- coding_variants$refsnp_id[!is.na(coding_variants$refsnp_id)]
```

# Retrieve SIFT annotations from Ensembl
```{r}
sift_annotations <- getBM(
  attributes = c(
    "refsnp_id",
    "ensembl_transcript_stable_id",   # transcript-level info
    "sift_prediction",                # predicted effect of amino acid change
    "sift_score"                      # probability score (0–1)
  ),
  filters = "snp_filter",
  values = coding_rsids,
  mart = snp_mart
)
```


```{r}
# Optional: reload previously saved annotations instead of querying BioMart
#sift_annotations_connection <- url("https://raw.githubusercontent.com/ana123136/Variant-Annotation-Using-Bioconductor/main/sift_annotations_backup.rds", "rb")
#sift_annotations <- readRDS(sift_annotations_connection)
#close(sift_annotations_connection)
```

# Keep all transcript-level SIFT info
```{r}
sift_long <- sift_annotations %>%
  filter(!is.na(sift_score) | !is.na(sift_prediction)) %>% # remove variants where no SIFT prediction is available
  rename(transcript_id = ensembl_transcript_stable_id)
```

# Join with coding variants
```{r}
coding_variants_long <- coding_variants %>%
  left_join(sift_long, by = "refsnp_id", relationship = "many-to-many")
```

**Histogram of distribution of SIFT scores**

# Histogram of SIFT scores
```{r warning=FALSE}
ggplot(coding_variants_long, aes(x = sift_score)) +
  geom_histogram(binwidth = 0.05, fill = "deeppink", color = "black") +
  
  # Create dummy rectangles to generate a legend
  geom_rect(aes(xmin = 0, xmax = 0.05, ymin = 0, ymax = 0, fill = "0–0.04: Deleterious"), alpha = 0) +
  geom_rect(aes(xmin = 0.05, xmax = 1, ymin = 0, ymax = 0, fill = "0.05–1: Tolerated"), alpha = 0) +
  
  scale_fill_manual(
    name = "SIFT Score Interpretation",
    values = c("0–0.04: Deleterious", "0.05–1: Tolerated")
  ) +
  
  labs(
    title = "Distribution of SIFT Scores",
    x = "SIFT Score",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

**Bar chart of SIFT predictions**

# Count SIFT predictions
```{r}
sift_counts <- coding_variants_long %>%
  filter(!is.na(sift_prediction) & sift_prediction != "") %>%
  count(sift_prediction)
```

# Plot of SIFT predictions for coding variants
```{r}
ggplot(sift_counts, aes(x = reorder(sift_prediction, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(
    low = "#FFC0CB",   # light pink
    high = "#FF1493",  # deep pink
    name = "Variant Count"
  ) +
  labs(
    title = "SIFT Predictions for Coding Variants",
    x = "SIFT Prediction",
    y = "Variant Count"
  ) +
  theme_minimal()
```
#"low-confidence" indicates that prediction is uncertain due to limited alignment evidence

# Check variant rs528975748, which has more than one SIFT prediction
```{r}
multiple_sift_predictions <- coding_variants_long %>% filter(refsnp_id == "rs528975748")
print(multiple_sift_predictions)
#https://www.ensembl.org/Homo_sapiens/Variation/Mappings?db=core;r=19:45238320-45239320;v=rs528975748;vdb=variation;vf=1026460548
```

## Variant annotation – summary
# Filter variants with SIFT <= 0.4 and meaningful phenotypes
```{r}
filtered_variants <- coding_variants_long %>%
  filter(
    !is.na(sift_score) & sift_score <= 0.4,
    !is.na(phenotype_description),
    phenotype_description != "",
    phenotype_description != "ClinVar: phenotype not specified"
  ) %>%
  select(
    refsnp_id,
    consequence_type_tv,
    phenotype_description,
    transcript_id,
    sift_prediction,
    sift_score
  )
```

# Print the resulting table
```{r}
print(filtered_variants)
```
# This table represents a set of potentially deleterious coding variants with meaningful phenotypic associations
# It can be used for:
# 1) further experimental validation
# 2) exploring links between coding changes and complex traits or diseases
